---
title: "Case Study 2, Pt. 1"
author: "Nathaniel Brown, In Hee Ho, Sarah Zimmermann"
date: "September 28, 2017"
output: html_document
---
```{r}
library(ggplot)
library(survival)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

dat <- read.table("kellydat.txt", header=T)
```

#Introduction

The purpose of this project is to investigate the effect of the number of symptoms, gender, and race, on time to neurological assessment. 

#Exploratory Data Analysis
```{r}
missing_d <- (which(is.na(dat)))
dat[missing_d,] # no NA's

dat$race = "Others"
dat$race[dat$black==1]="Black"
dat$race[dat$hisp==1]="Hispanic"
dat$gender = "male"
dat$gender[dat$male==0]="female"
dat$symptom = 0
dat$symptom[dat$sn1==1]=1
dat$symptom[dat$sn2==1]=2
dat$symptom[dat$sn3==1]=3
dat$symptom[dat$all4==1]=4

ggplot(data = dat, aes(x=race,y=nctdel)) + geom_boxplot(aes(fill=gender))
ggplot(data = dat, aes(x=as.factor(symptom), y=nctdel)) +geom_boxplot()
ggplot(data = dat, aes(x=race, y=symptom)) + geom_boxplot(aes(fill=gender))
```


#Analysis and Approach

```{r}

```

http://influentialpoints.com/Training/coxs_proportional_hazards_regression_model-principles-properties-assumptions.htm#modmch

These are the assumptions for Cox regression according to the link: 

1.    The mechanisms giving rise to censoring of individuals must not be related to the probability of an event occurring. This assumption is known as non-informative censoring and applies for nearly all forms of survival analysis.

2.    Survival curves for different strata must have hazard functions that are proportional over time. This is the proportional hazards assumption discussed above for a single explanatory variable and for the multivariate model.

3.    There should be a linear relationship between the log hazard and each covariate. This is best checked using residual plots
